Limited plasticity but increased variance in physiological rates across ectotherm populations under climate change

### 2024-07-15 ms.qmd
Fixing up responses! I nee dto add line numbers at the end which I think will be easy.Added a few more refs and also re-rendered for another read!

Climate change is expected to result in warmer and more variable thermal environments globally [@Suarez2023; @Easterling2000; @Ummenhofer2017]. Greater thermal variability is predicted to pose strong selection pressure that leads to genetic adaptation and/or the evolution of adaptive phenotypic plasticity -- both of which are considered important for population resilience to human-induced climate change [@Seebacher2015-wi; @Merila2014; @Chevin2015; @Chevin2017; @Chevin2010; @Nunney2016; @seebacher2023can; @cooke2021one]. Without plasticity or adaptation, high extinction rates are expected unless organisms can migrate to track suitable habitats [@Cahill2012; @Nunney2016].

Reversible phenotypic plasticity, such as physiological acclimation, is relatively rapid and can be fine-tuned to environmental conditions making it the first ‘line-of-defense’ against environmental change [@Dewitt1998; @Scheiner2020]. For example, physiological rates are known to speed up as temperature increases because of the thermodynamic effects on chemical reaction rates -- so called 'acute' temperature responses (@fig-concept). However, longer-lasting (days-weeks) temperature increases move environmental conditions away from thermal optima of physiological rate functions (i.e., thermal performance curves) can be mitigated by acclimation, which adjust reaction rates [@Seebacher2015-wi; @Havird2020-mj]. Physiological acclimation is driven by endocrine and epigenetic processes that change the underlying physiology to allow organisms to maintain physiological performance around a fitness optimum despite changes in the environment [@seebacher2019histone; @little2013thyroid; @taff2016endocrine]. Acclimation therefore alters acute thermal sensitivity to offset the potentially negative effects of acute temperature changes (e.g., higher energetic demands). Acclimation, however, does not normally result in physiological rates returning to their original state (‘complete’ compensation - *sensu* @huey1999testing), but rather are only partially compensated [@huey1999testing; @Havird2020-mj].

Acclimation is expected to evolve in populations when environmental variability is high but predictable and the costs of plasticity are low [@Chevin2017; @Scheiner2020; @Nunney2016; @Dewitt1998; @Reed2010]. Thermal variation and predictability are known to differ across habitats (terrestrial, marine and freshwater) [@steele2019comparison], and as such, we expect organisms within these habitats to vary in their capacity for plastic adjustments in physiology. Species occupying terrestrial habitats have been shown to exhibit weak acclimation abilities, and so, are likely to be particularly vulnerable to climate change given their greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @hoffmann2013upper; @morley2019physiological]. In contrast, marine and freshwater habitats appear to have greater physiological acclimation capacity [e.g., @Seebacher2015-wi; @Pottier2022]. However, the focus of research up to now has been primarily on mean physiological responses neglecting how variability in physiological rates might also be impacted at higher temperatures.

As mean physiological rates increase with temperature there is reason to believe that the intrapopulation variability might also be impacted. Positive mean-variance relationships are common across biology suggesting that, as rates increase, so to should variation in rates (i.e., Taylor's Law; []).

@@

We show that biochemical traits (e.g., enzymes) were more likely to acclimate compared to whole-organisms traits as predicted, but only for marine ectotherms. In contrast, both whole-organism and biochemical traits were equally likely to acclimate in freshwater ectotherms. Unsurprisingly, we observed no distinction between traits in terrestrial ectotherms......FRANK?

@@

Despite these expectations, more broadly, we show little difference between early and late life acclimation capacities and little change in the variance in physiological rates across habitats. This may not be too surprising given that such responses are likely context or trait dependent [@carter2020life; @moghadam2019heat]. The lack of differences we observed may also be due to the fact that both juvenile and adult animals occupy similar thermal niches, disperse to a similar extent and exhibit comparable thermoregulatory behaviors making physiological responses to temperature similar. It is possible that temperatures experienced earlier in development (e.g., egg or larval stages) exhibit greater capacity for acclimation because of limited dispersal capabilities and very different habitats. Some empirical evidence supports this in insects. For example, @moghadam2019heat showed that larval *Drosophila* were more plastic in their heat hardening responses compared to adults. A focus on collecting more detailed information on behaviour, dispersal and thermal environments experienced by different life stages is likely to provide a more complete picture on when plasticity differs. To date, however, there are few studies that have focused on how variability in responses change. We would also encourage more empirical focus on this questions given the potential ecological and evolutionary implications.

@@

Enhanced knowledge of how variation in physiological rates vary across populations and species, and the degree to which they can be adjusted in response to the environment, may lead to more informed predictions about the ecological and evolutionary dynamics of natural populations [@forsman2015rethinking; @seebacher2023can; @cooke2021one; @Sanderson2023]. While we show general patterns across taxa and habitats it is important to recognise that this does not mean such patterns apply to all populations. Substantial variation in acclimation responses and changes in variance exist among populations and traits, as evidenced by wide prediction intervals and substantial study- and trait-level variance estimates. Nonetheless, quantitative measures on the changes in variance in physiological rates could be better incorporated into physiological and ecological models to provide more nuanced, and possibly more realistic, predictions about the impacts of climate change on natural populations. While we do not yet understand the relative contribution of environmental and genetic factors to variance changes, models could better decouple how different levels of heritability and total variance impact evolutionary and ecological predictions. Our meta-analysis now provides the opportunity to parameterise such models, and ensure they are better aligned with empirical findings.

>>>>

Climate change is expected to result in warmer and more variable thermal environments globally [@Suarez2023; @Easterling2000; @Ummenhofer2017]. Greater thermal variability is predicted to pose a strong selection pressure that leads to genetic adaptation and/or the evolution of adaptive phenotypic plasticity -- both of which are considered important for population resilience to human-induced climate change [@Seebacher2015-wi; @Merila2014; @Chevin2015; @Chevin2017; @Chevin2010; @Nunney2016; @seebacher2023can; @cooke2021one]. Without plasticity or adaptation, high extinction rates are expected unless organisms can migrate to track suitable habitats [@Cahill2012; @Nunney2016].

Reversible phenotypic plasticity, such as physiological acclimation, is relatively rapid and can be fine-tuned to environmental conditions making it the first ‘line-of-defense’ against environmental change [@Dewitt1998; @Scheiner2020]. For example, physiological rates are known to speed up as temperature increases because of the thermodynamic effects on chemical reaction rates -- so called 'acute' temperature responses (@fig-concept). However, longer-lasting (days-weeks) temperature increases that move environmental conditions away from thermal optima of physiological rate functions (i.e., thermal performance curves) can be mitigated by acclimation, which adjust reaction rates [@Seebacher2015-wi; @Havird2020-mj]. Physiological acclimation is driven by endocrine and epigenetic processes that change the underlying physiology to allow organisms to maintain physiological performance around a fitness optimum despite changes in the environment [@seebacher2019histone; @little2013thyroid; @taff2016endocrine]. Acclimation therefore alters acute thermal sensitivity to offset the potentially negative effects of acute temperature changes (e.g., higher energetic demands). Acclimation, however, does not normally result in physiological rates returning to their original state (‘complete’ compensation - *sensu* @huey1999testing), but rather are only partially compensated [@huey1999testing; @Havird2020-mj].

Acclimation is expected to evolve in populations when environmental variability is high but predictable and the costs of plasticity are low [@Chevin2017; @Scheiner2020; @Nunney2016; @Dewitt1998; @Reed2010]. Thermal variation and predictability are known to differ across habitats (terrestrial, marine and freshwater) [@steele2019comparison], and as such, we expect organisms within these habitats to vary in their capacity for plastic adjustments in physiology. Species occupying terrestrial habitats have been shown to exhibit weak acclimation abilities, and so, are likely to be particularly vulnerable to climate change given their greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @hoffmann2013upper; @morley2019physiological]. In contrast, marine and freshwater habitats appear to have greater physiological acclimation capacity [e.g., @Seebacher2015-wi; @Pottier2022]. However, the focus of research up to now has been primarily on mean physiological responses neglecting how variability in physiological processes might also be impacted by higher temperatures.

As mean physiological rates increase with temperature there is reason to believe that the intrapopulation variability will also be impacted. Positive mean-variance relationships are common across biology suggesting that, as physiological rates increase with temperature, so to should variation (i.e., Taylor's Law; [@giometto2015sample]). As biochemical reaction rates increase with temperatures

@@

We show that biochemical traits (e.g., enzyme activities) were more likely to acclimate compared to whole-organisms traits as predicted, but only for marine ectotherms. In contrast, both whole-organism and biochemical traits were equally likely to acclimate in freshwater ectotherms. Unsurprisingly, we observed no distinction between traits in terrestrial ectotherms......FRANK?

@@

Despite these expectations, more broadly, we show little difference between early and late life acclimation capacities and little change in the variance in physiological rates across habitats. This may not be too surprising given that such responses are likely context or trait dependent [@carter2020life; @moghadam2019heat]. The lack of differences we observed may be because both juvenile and adult animals occupy similar thermal niches, disperse to a similar extent and exhibit comparable thermoregulatory behaviors making physiological responses to temperature similar. It is possible that temperatures experienced earlier in development (e.g., egg or larval stages) exhibit greater capacity for acclimation because of limited dispersal capabilities and very different habitats. Some empirical evidence supports this in insects. For example, @moghadam2019heat showed that larval *Drosophila* were more plastic in their heat hardening responses compared to adults. A focus on collecting more detailed information on behaviour, dispersal and thermal environments experienced by different life stages is likely to provide a more complete picture on when plasticity differs. To date, however, there are few studies that have focused on how variability in responses change. We would also encourage more empirical focus on this questions given the potential ecological and evolutionary implications.

@@

Enhanced knowledge of how variation in physiological rates vary across populations and species, and the degree to which they can be adjusted in response to the environment leads to more informed predictions about the ecological and evolutionary dynamics of natural populations [@forsman2015rethinking; @seebacher2023can; @cooke2021one; @Sanderson2023]. We show general patterns across taxa and habitats that provide a foundation to understand the relationship between plasticity and trait variance, as well as particular trade-offs that could impact the benefits (or lack thereof) of acclimation. It is important to recognise, however, that these patterns do not necessarily apply to all populations. Substantial variation in acclimation responses and changes in variance exist among populations and traits, as evidenced by wide prediction intervals and substantial study- and trait-level variance estimates. Conservation efforts are often targeted at particular populations or species, and taxonomic differences are important in this context. Regardless, quantitative measures of the changes in variance in physiological rates could be better incorporated into physiological and ecological models to provide more nuanced, and possibly more realistic, predictions about the impacts of climate change on natural populations. While we do not yet understand the relative contribution of environmental and genetic factors to variance changes, models could better decouple how different levels of heritability and total variance impact evolutionary and ecological predictions. Our meta-analysis now provides the opportunity to parameterise such models, and ensure they are better aligned with empirical findings.
